A Decision Procedure for (Co)datatypes in SMT Solvers

### 2015-02-17 tex/body.tex
more

\begin{conf}
Detailed proofs of soundness and completeness are included in the technical report
associated with this paper \cite{our-report}.
\end{conf}%
The procedure is implemented in the SMT solver CVC4 as a combination
of rewriting and a theory solver
(Section~\ref{sec:the-theory-solver}).
It consists of about 2000 lines of \cpp{} code, among which \NOK{1600} are
shared between datatypes and codatatypes. The code is integrated in the
development version of the solver and is expected to be part of the CVC4~1.5 release.
%
An evaluation on %hand-crafted examples and on
problems generated from Isabelle theories using the Sledgehammer tool
demonstrates the usefulness of the approach (Section~\ref{sec:experimental-results}).

@@

\begin{rep}\paragraph{Related Work.} Although\end{rep}%
\begin{conf}Although\end{conf} it was written a few years ago,
Barrett et al.\ \cite{barrett-et-al-2007} still provides a good account of
related work. Since then, datatypes have been added to the SMT solver
Z3 in unpublished work \begin{rep}by Leonardo de Moura \end{rep}and to a SPASS-like %prototypical
superposition prover \begin{rep}by Daniel Wand\end{rep} \cite{wand-2014}.
Closely related is the automatic structural
induction in superposition provers \cite{kersani-peltier-2013,wand-weidenbach-201x}
and SMT solvers \cite{reynolds-kuncak-2015}.
%
Datatypes are explicitly supported by the higher-order model finder Refute
\cite{weber-2008}. Its successor, Nitpick \cite{blanchette-nipkow-2010}, can
also find models involving cyclic codatatype values. The $\mu$-notation is
inspired by the $\mu$-calculus \cite[etc.]{kozen-1983,endrullis-et-al-2011}.

@@

\newcommand\gandl{G\&L}
\newcommand\HD[1]{\hbox to2.3em{\hfill\!\!\!\!#1\!\!\!\!\hfill}}

@@

CVC4 was run on each problem several times, with the support for datatypes and
codatatypes either enabled or disabled. The contributions of the acyclicity and
uniqueness rules were also measured, so find out whether these somewhat
expensive rules are useful in practice. Even when the decision procedure is
disabled, some of the generated axioms typically include basic properties of
constructors and selectors, which the decision procedure would recognize as
tautologies.
%This allows us to answer the question, \relax{What are the
%benefits of activating the decision procedure as opposed to letting
%Sledgehammer do what it would normally do?}

@@

\begin{table*}[tbh!]
\normalsize
\begin{center}\begin{tabular}{@{\;}l@{\kern1.5em}cc@{\kern.5em}cc@{\kern.5em}cc@{\kern1.3em}cc@{\;}}
  & \multicolumn{2}{c}{Distro} & \multicolumn{2}{c}{AFP} & \multicolumn{2}{c}{\gandl} & \multicolumn{2}{c}{All} \\
  & \HD{Z3} & \HD{CVC4} & \HD{Z3} & \HD{CVC4} & \HD{Z3} & \HD{CVC4} & \HD{Z3} & \HD{CVC4}
\MIDRULE
No (co)datatypes
  & 209 & 221 & 777 & 785 & \phantom{0}51 & \phantom{0}52 & 1037 & 1048 \\[\jot]
%Datatype simplication only
%  &  \\
Datatypes w/o \rn{Acyclic}
  & & 235 & & 770 & & \phantom{0}61 & & 1066 \\
Full datatypes
  & 213 & 227 & 791 & 786 & \phantom{0}51 & \phantom{0}52 & 1055 & 1065 \\[\jot]
%Codatatype simplication only
%  &  \\
Codatatypes w/o \rn{Unique}
  & & 230 & & 794 & & \phantom{0}65 & & 1089 \\
Full codatatypes
  & & 231 & & 794 & & \bfseries \phantom{0}68 & & 1093 \\[\jot]
Full (co)datatypes
  & & \bfseries 237 & & \bfseries 802 & & \bfseries \phantom{0}68 & & \bfseries 1107% \\[\jot]
\end{tabular}\end{center}
\caption{\,Number of solved goals for the three benchmark suites}
\label{tab:bench}
\end{table*}

>>>>

\begin{conf}
Detailed proofs of soundness and completeness are included in the technical report
associated with this paper \cite{our-report}.
\end{conf}%
The procedure is implemented in the SMT solver CVC4 \cite{barrett-et-al-2011} as a combination
of rewriting and a theory solver
(Section~\ref{sec:the-theory-solver}).
It consists of about 2000 lines of \cpp{} code, among which \NOK{1600} are
shared between datatypes and codatatypes. The code is integrated in the
development version of the solver and is expected to be part of the CVC4~1.5 release.
%
An evaluation on %hand-crafted examples and on
problems generated from Isabelle theories using the Sledgehammer tool
demonstrates the usefulness of the approach (Section~\ref{sec:experimental-results}).

@@

\begin{rep}\paragraph{Related Work.} Although\end{rep}%
\begin{conf}Although\end{conf} it was written a few years ago,
Barrett et al.\ \cite{barrett-et-al-2007} still provides a good account of
related work. Since then, datatypes have been added to the SMT solver
Z3 \cite{de-moura-bjoerner-2008} in unpublished work \begin{rep}by Leonardo de Moura \end{rep}and to a SPASS-like %prototypical
superposition prover \begin{rep}by Daniel Wand\end{rep} \cite{wand-2014}.
Closely related is the automatic structural
induction in superposition provers \cite{kersani-peltier-2013,wand-weidenbach-201x}
and SMT solvers \cite{reynolds-kuncak-2015}.
%
Datatypes are explicitly supported by the higher-order model finder Refute
\cite{weber-2008}. Its successor, Nitpick \cite{blanchette-nipkow-2010}, can
also find models involving cyclic codatatype values. The $\mu$-notation is
inspired by the $\mu$-calculus \cite[etc.]{kozen-1983,endrullis-et-al-2011}.

@@

\newcommand\gandl{G\&L}
\newcommand\HD[1]{\hbox to2.25em{\footnotesize\hfill\!\!\!\!#1\!\!\!\!\hfill}}

@@

CVC4 was run on each problem several times, with the support for datatypes and
codatatypes either enabled or disabled. The contributions of the acyclicity and
uniqueness rules were also measured, so find out whether these somewhat
expensive rules are useful in practice. Even when the decision procedure is
disabled, some of the generated axioms typically include basic properties of
constructors and selectors, which the decision procedure would recognize as
tautologies. Z3 is included for comparison when this is possible.
%This allows us to answer the question, \relax{What are the
%benefits of activating the decision procedure as opposed to letting
%Sledgehammer do what it would normally do?}

@@

\begin{table*}[t!]
\normalsize
\begin{center}\begin{tabular}{@{\;}l@{\kern1.5em}cc@{\kern.5em}cc@{\kern.5em}cc@{\kern1.3em}cc@{\;}}
  & \multicolumn{2}{c}{Distro} & \multicolumn{2}{c}{AFP} & \multicolumn{2}{c}{\gandl} & \multicolumn{2}{c}{All} \\
  & \HD{Z3} & \HD{CVC4} & \HD{Z3} & \HD{CVC4} & \HD{Z3} & \HD{CVC4} & \HD{Z3} & \HD{CVC4}
\MIDRULE
No (co)datatypes
  & 209 & 221 & 777 & 785 & \phantom{0}51 & \phantom{0}52 & 1037 & 1048 \\[\jot]
%Datatype simplication only
%  &  \\
Datatypes w/o \rn{Acyclic}
  & & 235 & & 770 & & \phantom{0}61 & & 1066 \\
Full datatypes
  & 213 & 227 & 791 & 786 & \phantom{0}51 & \phantom{0}52 & 1055 & 1065 \\[\jot]
%Codatatype simplication only
%  &  \\
Codatatypes w/o \rn{Unique}
  & & 230 & & 794 & & \phantom{0}65 & & 1089 \\
Full codatatypes
  & & 231 & & 794 & & \bfseries \phantom{0}68 & & 1093 \\[\jot]
Full (co)datatypes
  & & \bfseries 237 & & \bfseries 802 & & \bfseries \phantom{0}68 & & \bfseries 1107% \\[\jot]
\end{tabular}\end{center}
\caption{\,Number of solved goals for the three benchmark suites}
\label{tab:bench}
\end{table*}
